Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-16 demystify.tex
updates

\subsection{Common Pitfalls}

@@

\paragraph{Overexaggeration Lors} Many GRFP applicants are from top schools with well-known professors that are very good at writing. The issue is that they often use flowery language and overrate their students. Reviewers are often professors themselves and can detect B.S. So you should ask your LoR writers to be honest, tone down the language (similar to overclaiming research results when writing papers), and be specific with concrete examples to demonstrate why you're ``the best''.

@@

\paragraph{Redacted info.} Unlike the NSF GRFP, for NDSEG reviewers \emph{do not} see your LoRs, i.e., we do not know who wrote them or what they wrote. Moreover, we only see \emph{redacted versions} of your submitted materials (applicants are asked to provide full and redacted versions of your research plan, CV, and other materials). Thus we do not know what schools your went to, where you did your internships, your publications or research experience.

Thus, reviewers mainly get to know you and evaluate your application by reading your research plan and personal essays. This is for \emph{fairness and avoiding bias} based on the school you went to, places you published at, your LoR writers, etc.  Nonethless, I felt I am missing too much information. However, it could be that external reviewers like myself help with the first cut based on the redacted materials, and then there are internal reviewers who makes the final decision based on the full materials.

@@

\begin{commentbox}[Vu:]
I find a bit difficult to evaluate NDSEG applications because I do not have the full picture.  For example, a student mentioned that they have published in the topic they proposed but redacted all information about the publication, so I have no idea where they published, what the paper is about, etc.  
It is also difficult to evaluate the academic background of the student, e.g., some student just started grad school and so their transcript is essentially blank, and I do not know where they went to undergrad, etc. It is also annoying to see CVs with so many blanked out information. They might as well should just submit a blank CV one. 
\\

In the beginning I find it surprising that I don't get to see the LoRs, but I understand the reasoning behind it and gradually think it is a good idea. As mentioned above, LoR writers, especially those from top schools, often overrate their students and thus can bias the evaluation. So I don't find I am missing much by not seeing the LoRs.
\end{commentbox}

\subsection{Common Pitfalls and Tips}
Based on my experience in reviewing, here are some common pitfalls to avoid:
\begin{itemize}

@@

%- use latex,  do not write in word and especially use left justification  (see how CS and engineer people publish papers)  
\item If you have an advisor or Lor writer, ask them to look at your writing/essays. This is because you are new to the field and research writing, and your ideas might not be accurate or have already been explored.

@@

>>>>

\subsection{Common Pitfalls}\label{sec:grfp-pitfalls}

@@

\paragraph{Overexaggeration Lors} Many GRFP applicants are from top schools with well-known professors that are very good at writing. The issue is that they often use flowery language and completely overrate their students. Reviewers are often professors themselves and can sense this overexaggeration.  Moreover, they often know the LoR writers and can compare the LoR with the student's application and see if they match.   So you should ask your LoR writers to be honest, tone down the language (similar to overclaiming research conributions when writing papers), and be specific with concrete examples to demonstrate why you're \emph{``the best''}.

@@

\paragraph{Redacted info.} Unlike the NSF GRFP, for NDSEG reviewers \emph{do not} see your LoRs, i.e., we do not know who wrote them or what they wrote. Moreover, we only see \emph{redacted versions} of your submitted materials (applicants are asked to provide full and redacted versions of your research plan, CV, and other materials). Thus, we do not know what schools your went to, where you did your internships, your publications or research experience.

Reviewers mainly get to know you and evaluate your application by reading your research plan and personal essays. This is for \emph{fairness and avoiding bias} based on the school you went to, places you published at, your LoR writers, etc.  Nonethless, I felt I am missing too much information. However, it could be that external reviewers like myself help with the first cut based on the redacted materials, and then there are internal reviewers who makes the final decision based on the full materials.

@@

\begin{commentbox}[Vu:]
I find a bit difficult to evaluate NDSEG applications because I do not have the full picture.  For example, a student mentioned that they have published in the topic they proposed but redacted all information about the publication, so I have no idea where they published, what the paper is about, etc.  
It is also difficult to evaluate the academic background of the student, e.g., some student just started grad school and so their transcript is essentially blank, so I have no idea on their academic background.  It is also annoying to see CVs with so many blanked out information. They might as well should just submit a blank CV one. 
\\

In the beginning I find it surprising that I don't get to see the LoRs, but I understand the reasoning behind it and gradually think it is a good idea. As mentioned above, LoR writers for NSF GRFP (\autoref{sec:grfp-pitfalls}), especially those from top schools, often overrate their students and thus can bias the evaluation. So I don't find I am missing much by not seeing the LoRs.
\end{commentbox}

\subsection{Common Pitfalls and Tips}\label{sec:ndseg-pitfalls}

These are some common pitfalls I see in NDSEG applications and some tips to avoid them.  While some of these are similar to NSF GRFP (\autoref{sec:grfp-pitfalls}), some are specific to NDSEG.

@@

%- use latex,  do not write in word and especially use left justification  (see how CS and engineer people publish papers)

@@

Finally, just as with the NSF GRFP and even PhD SOPs, ask your LoR writers and adviser to look at your writing/essays. You are new to the field and research writing, and your ideas might not be accurate or have already been explored. It's better that they tell you now than for you to find out later from reviewers.
